---
title: "Friedman: Package Documentation"
author: "Animesh Singh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Friedman: Package Documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 4.5
)
devtools::load_all()
```

## Overview

This package started as a way to avoid rewriting the same ggplot2 setup every
time I start a new dataset. It has six plotting functions and one summary
function, all built around a custom S3 class called `friedman_plot` that
keeps the plot, the data, and some metadata together in one object.

---

## Dataset: `ai_job_impact`

The package ships with `ai_job_impact`, a 2,000-row survey dataset on how AI
adoption has affected employees across industries.

```{r}
data(ai_job_impact)
dim(ai_job_impact)
head(ai_job_impact[, 1:8])
```

---

## The `friedman_plot` S3 Class

Every plotting function returns a `friedman_plot` object, which is just a
named list with these fields:

- `$plot` — the ggplot2 object
- `$data` — the data frame that was passed in
- `$type` — what kind of plot it is (e.g. `"histogram"`, `"scatter"`)
- `$vars` — which columns were used
- `$nrow` / `$ncol` — dimensions of the data

Three methods are registered for it:

- `print()` — prints a one-liner and renders the plot
- `summary()` — shows metadata and basic stats for the variables used
- `plot()` — re-renders the plot if you need it again

```{r}
p <- quick_hist(ai_job_impact, Age)
class(p)
names(p)
```

```{r}
summary(p)
```

---

## Functions

### `quick_hist(data, col, bins = 30)`

A ggplot2 histogram for any numeric column. Column is passed unquoted. The
title and axis label are derived automatically from the column name.

```{r}
quick_hist(ai_job_impact, Age)
```

```{r}
quick_hist(ai_job_impact, Salary_Before_AI, bins = 20)
```

---

### `quick_scatter(data, x, y)`

Scatter plot of two numeric columns with a linear trend line (`geom_smooth(method = "lm")`).
Both columns are passed unquoted.

```{r}
quick_scatter(ai_job_impact, Years_Experience, Salary_After_AI)
```

---

### `quick_box(data, x, y)`

Box-and-whisker plot of a numeric column grouped by a categorical variable.
Both arguments are strings. The grouping variable is coerced to a factor
internally.

```{r}
quick_box(ai_job_impact, "Job_Status", "Salary_After_AI")
```

---

### `quick_cor(data, method = "pearson")`

Computes pairwise correlations for all numeric columns and renders a
colour-coded heatmap annotated with coefficient values. Non-numeric columns
are silently dropped. The `method` argument accepts `"pearson"`,
`"kendall"`, or `"spearman"`.

This function is more involved than the basic wrappers — it builds an
intermediate tidy data frame from the correlation matrix before plotting,
which allows `geom_tile` to map the full matrix correctly.

```{r fig.width=8, fig.height=6}
num_cols <- c("Age", "Years_Experience", "Salary_Before_AI",
              "Salary_After_AI", "Work_Hours_Per_Week", "Job_Satisfaction")
quick_cor(ai_job_impact[num_cols])
```

---

### `quick_pct_change(data, before_col, after_col, group_col)`

Computes the median percentage change `(after - before) / before * 100` for
each group and plots the results as a horizontal diverging bar chart. Bars
are coloured blue for positive change and salmon for negative, with a dashed
zero line for reference. Works on any two numeric before/after columns —
salaries, test scores, prices, etc.

```{r}
quick_pct_change(ai_job_impact,
                 "Salary_Before_AI", "Salary_After_AI", "Job_Status")
```

```{r}
quick_pct_change(ai_job_impact,
                 "Salary_Before_AI", "Salary_After_AI", "AI_Adoption_Level")
```

---

### `quick_summary(data)`

Prints a tidy summary table of all numeric columns: mean, median, standard
deviation, min, and max. Returns the table invisibly as a data frame.

```{r}
quick_summary(ai_job_impact[c("Age", "Salary_Before_AI", "Salary_After_AI",
                               "Job_Satisfaction")])
```

---

### `quick_compare(data, num_col, group_col)`

Overlaid density curves for a numeric column split by a grouping variable.
Useful for comparing distributions across categories without reducing them
to box plot summaries.

```{r}
quick_compare(ai_job_impact, "Salary_After_AI", "Job_Status")
```

```{r}
quick_compare(ai_job_impact, "Job_Satisfaction", "AI_Adoption_Level")
```

---

## How it works under the hood

The three basic functions (`quick_hist`, `quick_scatter`, `quick_box`) take
unquoted column names, which I handle with `deparse(substitute())`. Inside
the `aes()` calls I use the `.data` pronoun from `rlang` — without it, R CMD
check throws notes about undefined variables, which I found out the hard way.

I added a small internal helper `.check_col()` that runs before every plot.
It checks that the column exists and is numeric when it needs to be, so you
get a clear error immediately instead of something cryptic from inside ggplot2.

I went with S3 over S4 because the object is just a named list.
S4 would add formal slot definitions and type validation which is overkill here.
The three methods (`print`, `summary`, `plot`) cover everything you'd want to
do with the output anyway.

---

## AI Use Disclosure

I used an AI assistant in a few limited areas of this project:

**Where it was used:** Formatting the roxygen documentation comments,
suggesting the correct syntax for `.data` pronoun usage inside `aes()` to
avoid R CMD check notes, helping structure the testthat test cases, and
unblocking a DESCRIPTION field error I couldn't track down (the redundant
`Author`/`Maintainer` fields conflicting with `Authors@R`).

**What it contributed:** Syntax fixes and documentation formatting. The
test structure was suggested by AI but I reviewed and adjusted each case to
match what the functions actually do.

**What I implemented independently:** The overall package concept and
function design, choice of which plots to include and how to parameterize
them, the decision to use S3 over S4, the `friedman_plot` class structure,
the `quick_cor` and `quick_pct_change` logic, dataset selection, and all
analysis and interpretation in this document.

---

## License

CC0 1.0 Universal — no rights reserved.
